---
title: "Introduction to R"
author: "Yong Jun"
date: "December 13, 2015"
output: 
  html_document:
    keep_md: true
---

## Why R?
- R & Python two of the most popular tools for data science work
- R is a language designed for the purpose of statistical programming and visualization
- It is a high level language
    - meaning it abstracts you away from many of the nitty gritty of details of doing statistical computation)
- It is open source
- Rich community of users that contribute libraries to do almost anything you need

## Difference with Python

#### Python

- General purpose programming language
- Statistical computation supported through packages like *NumPy*, *SciPy*, *pandas* etc
- Object-oriented

#### R

- Designed for statistical programming
- Has certain built-in design features that makes it suitable for this task
- Functional (through use of generic functions)
- R has [Hadley Wickham](http://priceonomics.com/hadley-wickham-the-man-who-revolutionized-r/)

## First, a bit on structured data

- Structured data: SQL tables
- Semi-structured: JSON, XML
- Unstructured: Plain text, Image, Audio
- Terminology:
    - Tables
    - Records: Row
    - Fields: Columns
- Each record has a **fixed set of fields**
- Values within a field must be of the **same data type**
    - Numeric
    - String
    - Date
    - Categorical
    - Binary

```{r, echo=FALSE}
data("mtcars")
mtcars
```

## What I'll be covering

There's so much to cover with regards to data science, machine learning or R but I don't have the time to go through all that. Plus there's also Python. So rather than taking a narrow focus of programming in R, I'll cover some of the special features in R that helps get you started in thinking about programming data science.

## Things about R
### Point 1: Everything is an Array

In R, the most basic data object is the 1 dimension array (vector). Try this:

Create an array:
```{r}
arr <- 1:10
arr
```

Array can be indexed or subsetted:
```{r, results='hold'}
arr[7]
arr[1:4]
arr[c(1, 3, 5)]
```

Array operations:
```{r, results='hold'}
a <- 1:5
b <- 6:10
a + b
b - a
a * b
a / 2
a^2
a %% 2 == 0
```

Vector can only have one data type:
```{r}
arr <- c(1, 'a', TRUE)
arr
class(arr)
```

Try creating a "scalar":
```{r, results='hold'}
scalar <- 9
scalar
scalar[1]
length(scalar)
```

Try adding two arrays of different length
```{r}
a <- 1:10
b <- 1:2
rbind(a, b, sum=a+b)
```

Mapping a function
```{r}
f <- function (n) {
  n * 2
}
f(a)
```

Reduce functions
```{r, results='hold'}
sum(a)
mean(a)
min(a)
max(a)
cumsum(a)
```

This feature of R abstract us away from manually doing element-wise operations on array. How array addition in usually carried out in languages without array operation (example Javascript):
```Javascript
var toReturn = []
for (var i = 0; i < a.length; i++) {
  toReturn.push(a[i] + b[i])
}
console.log(toReturn)
```

Not saying this cannot be done in other languages but additional libraries are needed (eg. Python -> Numpy, Javascript -> math.js). Matlab also builds on this feature but R brings it one step further by completely doing away with scalar data types. So everything is a vector.

Good read to understand the rational for vectorization in R:
[http://www.noamross.net/blog/2014/4/16/vectorization-in-r--why.html](http://www.noamross.net/blog/2014/4/16/vectorization-in-r--why.html)

### Point 2: R has data.frame

Vectors in R can be used to represent columns of data in a table. But a table is made up of a group of columns each by themselves must be of one data type but within themselves can have different data types. How do we address this?

Try using matrix:
```{r, results='hold'}
a <- 1:5
b <- c('a', 'b', 'c', 'd', 'e')
tb <- cbind(a, b)
tb
class(tb)
typeof(tb)
```

Not a good idea because matrix like array can have only one data types so the first column is converted to chracter type to match second column. The data structure in R that can support different data types is a **list**:
```{r}
tb <- list(a=a, b=b)
tb
```

But tables have another constraint. Every column must be the same length. How do we make sure of this? In R, the answer comes in the form of **data.frame**:
```{r, results='hold'}
tb <- data.frame(a, b)
tb
```

```{r, results='hold'}
x <- 1:2
print(try(tb <- data.frame(a, b, x)))
```

So **data.frame** is a special type of **list** in R that enforces same length for every member. Of course it doesn't end there, **data.frame** has other features and behaviors useful for table-form data analysis.

Subsetting a **data.frame**:
```{r, results='hold'}
typeof(mtcars)
class(mtcars)
mtcars[mtcars$mpg > 25, ]
```

Selecting specific columns:
```{r, results='hold'}
mtcars[mtcars$mpg > 25, 1:3]
mtcars[mtcars$mpg > 25, -(1:3)]
mtcars[mtcars$mpg > 25, c('mpg', 'hp', 'wt')]
```

**data.frame** can be passed into many functions in R to for further analysis. For example visualizing with **ggplot**
```{r}
library(ggplot2)
mtcars$cyl <- as.factor(mtcars$cyl)
gg <- ggplot(data=mtcars, aes(x=hp, y=mpg))
gg + geom_point(aes(color=cyl))
```

Python implements **data.frame** through the **pandas** library. Scientific plotting in Python is done using **matplotlib**.

### Point 3: Machine learning is made easy with R
Say we want to model 'Miles Per Gallon' (mpg) as a function of 'Horse Power' (hp) by plotting a best fit straight line (i.e. linear regression):
```{r}
fit <- lm(mpg~hp, data=mtcars)
summary(fit)
gg + geom_point(aes(color=cyl)) +
  geom_abline(intercept=fit$coefficients[1], slope=fit$coefficients[2])
```

Let's predict mpg of cars with hp of 100, 200 & 300:
```{r}
test <- data.frame(hp=c(100, 200, 300))
predicted_mpg <- predict(fit, test)
cbind(test, predicted_mpg)
```

How about adding transmission (auto/manual) as a predictor variable?
```{r}
mtcars$am <- as.factor(mtcars$am)
levels(mtcars$am) <- c('Auto', 'Manual')
fit <- lm(mpg~hp+am, data=mtcars)
summary(fit)
predicted_mpg <- predict(fit, mtcars)
ggplot(data=mtcars, aes(x=hp, y=mpg, color=am)) + geom_point()  +
  geom_line(aes(y=predicted_mpg))
```

Here, we demostrated a feature of R called **factor**. Which brings us to the next point

### Point 4: R has a special way to handle categorical data - **factor**

Factor converts an array of strings into integer representation
```{r}
data(iris)
head(iris)
summary(iris)
str(iris)
boxplot(Sepal.Length~Species, iris, col=c('red', 'dark orange', 'yellow'))
```

So you can see factor is an efficient way of storing categorical data. Factor is also useful for ordering and comparing nominal data (eg. First, Second, Third):
```{r}
orderedMonth <- c('Jan', 'Feb', 'Mar', 'Apr', 'May', 'Jun')
unorderedMonth <- c('May', 'Mar', 'Jan', 'Apr', 'Feb', 'Jun')
mon <- factor(unorderedMonth, levels=orderedMonth, ordered=TRUE)
mon
```

```{r, results='hold'}
mon[1] > mon[2]
mon[5] > mon[6]
```

Factor is also supported in Python through **pandas**

## What are some interesting things R can do?

### K-means clustering
```{r}
scaled_data <- scale(as.matrix(iris[,-5]))
km <- kmeans(scaled_data, 3)
iris$cluster <- as.factor(km$cluster)
ggplot(data=iris, aes(x=Sepal.Length, y=Sepal.Width, color=cluster, shape=Species)) +
  geom_point(size=5)
```

## Word Cloud
```{r}
library(wordcloud)
library(zipfR)
data("Brown.tfl")
Brown_subset <- Brown.tfl[order(Brown.tfl$f, decreasing=TRUE),][1:150,]
wordcloud(Brown_subset$type,
          Brown_subset$f,
          scale=c(7,1),
          color=brewer.pal(8,'Dark2'),
          random.order=FALSE,
          random.color=TRUE,
          rot.per=.3)
```

## Decision Tree
```{r}
library(rpart)
library(rattle)
data("swiss")
head(swiss)
summary(swiss)
swiss$Infant.Mortality <- ifelse(swiss$Infant.Mortality > 20, '> 20', '<= 20')
swiss$Infant.Mortality <- as.factor(swiss$Infant.Mortality)
fit <- rpart(Infant.Mortality~., data=swiss)
fancyRpartPlot(fit)
pred <- predict(fit, swiss, type='class')
cbind(Actual=swiss$Infant.Mortality, Predicted=pred)
```

## Interactive Chart
```{r}
library(reshape2)
GDP <- read.csv('resources/GDP.csv')
Population <- read.csv('resources/Population.csv')

GDP <- melt(GDP, id.vars='Year', variable.name='Country', value.name='GDP')
Population <- melt(Population, id.vars='Year', variable.name='Country', value.name='Population')

combined <- data.frame(Population, GDP)
combined$GDP <- combined$GDP * 1000000
combined$Population <- combined$Population * 1000
combined$Per_Capita <- combined$GDP / combined$Population
combined <- combined[!(combined$Country %in% c('Kuwait','UAE','Qatar')),]

gg <- function(year) {
  qplot(x=Per_Capita, y=GDP, size=log(Population,10), label=Country, data=combined[combined$Year==year,], log='y') + geom_text(hjust=-.5, size=3)
}
gg(1970)
```

```{r, eval=FALSE}
library(manipulate)
manipulate(gg(year), year=slider(min=1970, max=2008, step=2))
```

## Shiny App

- [Example 1](https://github.com/yongjun21/DevelopDataProducts)
- [Example 2](https://github.com/yongjun21/DataScienceCapstone)